---
title: "Introduction to R & RStudio"
author: "Jenna MacKinnon"
date: "2024-02-08"
output: html_document
---

***I acknowledge that I am speaking today from Kjipuktuk, and want to pay my respects to the original caretakers of this land, the Mi'kmaq.***

# 💃 **About Me**

**Email:** jenna.mackinnon\@nrc-cnrc.gc.ca

**Github:** yoitsjenna

-   Graphing & Data Visualization

-   Mapping

-   Data wrangling & tidying

-   Basic statistics

-   Multivariate statistics

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# 🖥️ The Basics

## R Studio Layout

-   Source -- Able to save and edit scripts continually 

-   Console -- interactive area where R commands are executed one at a time; it provides immediate feedback, allowing you to see the results of your code in real-time.

-   Environment -- current objects and data

-   History -- keeps track of the commands you have previously executed in the R console, including both individual commands and entire scripts

-   Packages -- list of some available packages and which ones are installed

-   Plots -- where plots are "printed"; also included in the R markdown script

-   Help -- help for functions, packages, and further learning

## **Mathematical Operations & Variables**

```{r}

a <- 8 + 8
a

100 - 50 

18 * 3

72 / 12

# The default to this function is the natural log (ln)
log(2.718)

# Rounds numbers with decimals -- default is a whole number
round(3.14159)

```

# 🛠️ Install and Load Packages

**Packages:** Extra functions, data, and/or compiled code designed to solve specific problems or perform certain tasks

```{r}

# install.packages("tidyverse")
library(tidyverse)

# install.packages("oce")
library(oce)

# install.packages("ocedata")
library(ocedata)

# install.packages("magrittr")
library(magrittr)

# https://cran.r-project.org/web/packages/spData/readme/README.html
# install.packages("spData")
library(spData)

# install.packages("extrafont")
library(extrafont)

# install.packages("hrbrthemes")
library(hrbrthemes)

# install.packages("plotly")
library(plotly)

# install.packages("heatmaply")
library(heatmaply)

# install.packages("virdis")
library(viridis)

# install.packages("circlize")
library(circlize)

# install.packages("devtools", dependencies = TRUE)
library(devtools)

# install.packages("patchwork")
library(patchwork)

# devtools::install_github("mattflor/chorddiag")
library(chorddiag) 

#install.packages("gapminder")
library(gapminder)

```

# 🔥 Creating a Heat Map

Source: <https://www.data-to-viz.com/graph/heatmap.html>

```{r}

# Load data 
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/multivariate.csv", header=T, sep=";")
colnames(data) <- gsub("\\.", " ", colnames(data))

# Select a few country
data <- data %>% 
  filter(Country %in% c("France", "Sweden", "Italy", "Spain", "England", "Portugal", "Greece", "Peru", "Chile", "Brazil", "Argentina", "Bolivia", "Venezuela", "Australia", "New Zealand", "Fiji", "China", "India", "Thailand", "Afghanistan", "Bangladesh", "United States of America", "Canada", "Burundi", "Angola", "Kenya", "Togo")) %>%
  arrange(Country) %>%
  mutate(Country = factor(Country, Country))

# Matrix format
mat <- data
rownames(mat) <- mat[,1]
mat <- mat %>% dplyr::select(-Country, -Group, -Continent)
mat <- as.matrix(mat)

p <- heatmaply(mat, 
        dendrogram = "none",
        xlab = "", ylab = "", 
        main = "",
        scale = "column",
        margins = c(60,100,40,20),
        grid_color = "white",
        grid_width = 0.00001,
        titleX = FALSE,
        hide_colorbar = TRUE,
        branches_lwd = 0.1,
        label_names = c("Country", "Feature:", "Value"),
        fontsize_row = 5, fontsize_col = 5,
        labCol = colnames(mat),
        labRow = rownames(mat),
        heatmap_layers = theme(axis.line=element_blank())
        )

p

```

# ♾️ Create a Chord Diagram

Source: <https://www.data-to-viz.com/graph/chord.html>

```{r}

# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/13_AdjacencyDirectedWeighted.csv", header=TRUE)

# short names
colnames(data) <- c("Africa", "East Asia", "Europe", "Latin Ame.",   "North Ame.",   "Oceania", "South Asia", "South East Asia", "Soviet Union", "West.Asia")
rownames(data) <- colnames(data)

# I need a long format
data_long <- data %>%
  rownames_to_column %>%
  gather(key = 'key', value = 'value', -rowname)

# parameters
circos.clear()
circos.par(start.degree = 90, gap.degree = 4, track.margin = c(-0.1, 0.1), points.overflow.warning = FALSE)
par(mar = rep(0, 4))

# color palette
mycolor <- viridis(10, alpha = 1, begin = 0, end = 1, option = "D")
mycolor <- mycolor[sample(1:10)]

# Base plot
chordDiagram(
  x = data_long, 
  grid.col = mycolor,
  transparency = 0.25,
  directional = 1,
  direction.type = c("arrows", "diffHeight"), 
  diffHeight  = -0.04,
  annotationTrack = "grid", 
  annotationTrackHeight = c(0.05, 0.1),
  link.arr.type = "big.arrow", 
  link.sort = TRUE, 
  link.largest.ontop = TRUE)

# Add text and axis
circos.trackPlotRegion(
  track.index = 1, 
  bg.border = NA, 
  panel.fun = function(x, y) {
    
    xlim = get.cell.meta.data("xlim")
    sector.index = get.cell.meta.data("sector.index")
    
    # Add names to the sector. 
    circos.text(
      x = mean(xlim), 
      y = 3.2, 
      labels = sector.index, 
      facing = "bending", 
      cex = 0.8
      )

    # Add graduation on axis
    circos.axis(
      h = "top", 
      major.at = seq(from = 0, to = xlim[2], by = ifelse(test = xlim[2]>10, yes = 2, no = 1)), 
      minor.ticks = 1, 
      major.tick.percentage = 0.5,
      labels.niceFacing = FALSE)
  }
)
```

```{r}
```

# 📈 Create an Interactive Bubble Plot

Source: <https://r-graph-gallery.com/interactive-charts.html>

```{r}

p <- gapminder %>%
  filter(year==1977) %>%
  ggplot( aes(gdpPercap, lifeExp, size = pop, color=continent)) +
  geom_point() +
  theme_bw()

ggplotly(p)
```

# 📚 Resources

-   [The R Graph Gallery -- Help and inspiration for R charts (r-graph-gallery.com](https://r-graph-gallery.com/index.html))
-   <http://www.cookbook-r.com/>
-   [LinkedIn Learning with Lynda: Online Training Courses for Creative, Technology, Business Skills](https://www.linkedin.com/learning/?trk=learning-serp_nav-header-logo&upsellOrderOrigin=default_guest_learning)
    -   Free access through some library memberships
-   **Tidyverse**
    -   <https://www.youtube.com/watch?v=8NnKhpG4JHA>
-   **R Cheatsheets**
    -   <https://www.rstudio.com/resources/cheatsheets/>
